---
title: "bitmexr 0.3.0: Place, modify and cancel your orders on BitMEX without leaving R! "
description: |
  bitmexr 0.3.0 brings some exciting new features to the package. bitmexr now            supports placing, editing and cancelling orders through BitMEX's API.
  In addition, the testnet version of the API is now supported soyou can try out         managing orders using the package in a risk free environment!
author:
  - name: Harry Fisher
    url: https://hfshr.netlify.com/
date: 05-25-2020
preview: update.jpg
repository_url: https://github.com/hfshr/distill_blog
draft: false
categories: 
  - bitmexr
  - R
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bitmexr)
```

# bitmexr 0.3.0 - Whats new?

The latest CRAN release of `bitmexr` supports placing, editing and cancelling orders on the BitMEX exchange. These actions are supported by the following functions:

- `place_order()`
- `edit_order()`
- `cancel_order` or  `cancel_all_orders()`

See the [package documentation](https://hfshr.github.io/bitmexr/reference/index.html) or the [BitMEX website](https://www.bitmex.com/api/explorer/) for details on what arguments can be used with each function.

# Practice on the Testnet

BitMEX has a fully functional Testnet exchange that lets you use a "practice" version of the live exchange. This is great for trying out a new trading idea or using trying these functions out in a risk free environment that is identical to the real thing.  Every function in `bitmexr` has a Testnet equivalent - simply use the "tn_" prefix to access the Testnet version of the function. For example:

```{r eval=FALSE}
place_order(symbol = "XBTUSD", price = 5000, orderQty = 100)
```

Will place an order on the live exchange, whereas:

```{r eval=FALSE}
tn_place_order(symbol = "XBTUSD", price = 5000, orderQty = 100)
```

Will place the order on the Testnet exchange.

# Authentication needed!

In order to use these new functions you must first obtain an API key and secret for the account you wish to use with `bitmexr`. With API key and secret in hand, you can add these to your ~/.Renviron file using:
```
bitmex_apikey = "api key"
bitmex_apisecret = "api secret"
```
for the live exchange or
```
testnet_bitmex_apikey = "testnet api key"
testnet_bitmex_apisecret = "tetsnet api secret"
```
for the Testnet exchange. 

See the [Authentication vignette](https://hfshr.github.io/bitmexr/articles/authentication.html) for more details.

# Extras

## Speed boost 

With authentication set up, `map_trades()` and `map_bucket_trades()` get a speed boost due to increased API limit when using authentication. Simply set `use_auth = TRUE` to get these functions running at 60 API call per minute `r icon::fa_rocket()`.

## Additional API endpoints

For additional API endpoints that don't have a dedicated API wrapper in `bitmexr`, it is possible use either:

- `get_bitmex()` for additional GET requests

and

- `post_bitmex()` for additional POST requests.

These functions just need the path to the API endpoint and a named list of valid parameters for that endpoint. For example, you could use:

```{r eval=FALSE}
chat <- get_bitmex(path = "/chat", 
                   args = list(channelID = 1, 
                              reverse = "true"))
```

To get the latest English trollbox chat data (not sure why anyone would want this, but it is possible!).

# What next?

Current ideas for a future release:

- Support for easily placing bulk orders 
- Rstudio addin for easy order submission

Of course, any feedback / ideas are more than welcome! [`r  icon::fa_github()`](https://www.github.com/hfshr/bitmexr)