---
title: "Building a python package: Reflections from an R user"
description: |
  In this post I note down some of my experiences with making my first python package, specifically highlighting some of the similarities and differences between R and python when it comes to package building. My hope is that R users looking to expand their pythonic horizons might find something useful!
author:
  - name: Harry Fisher
    url: https://hfshr.netlify.app/
date: 08-02-2020
preview: pkg.jpeg
output:
  distill::distill_article:
    self_contained: false
categories:
  - R
  - Python
repository_url: https://github.com/hfshr/distill_blog      
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
library(fs)
library(reticulate)

```

In an attempt to improve my python skills, I thought it would be fun to try and create a simple python package.
To ease myself into the world of python packages, I set out to implement a very simple version of my `bitmexr` R package - which I imaginatively called `bitmexpy`. 
For context, `bitmexr` is a simple API wrapper for www.bitmex.com that supports obtaining historical trade data as well as placing trades on the exchange. 
If interested, you can see `bitmexr` in action [here](https://hfshr.github.io/bitmexr/). 
For `bitmexpy`, I'm just going to stick to implementing the historical trade element, and leave the other functions for another day! 

# Package code

Central to any package is the code that facilitates what the package does. In R this is always included in the `R` subdirectory of the package directory. For example `bitmexr` has the following files: 

```{r eval=TRUE}
dir_tree("~/Downloads/ProjectR/bitmexr/R/")
```

This is fairly similar in python, but there are a few differences. 

* Python packages are essentially a collection of _modules_, with a module being a single namespace that can contain, functions, constants, class definitions and anything else. 
 
* Python scripts are usually located in a directory with the same name as the package - but this is not enforced, and scripts can also be included at the top level of the package directory.

* There can be subdirectories with modules within the main package directory, these can be considered subpackages.

* Each directory containing python scripts must contain an `__init__.py` file. This can be an empty file, but can also contain arbitrary code. The file gets run when the package is imported.

* You can use `import` statements in python package code (doing the equivalent `library()` call in an R package does not work, and instead a NAMESPACE file is used).

For example, here is the directory for the package code for `bitmexpy`:

```{r eval=TRUE}
dir_tree("~/Downloads/Python/bitmexpy/bitmexpy/")
```

Each of these files are modules within the package and can be imported separately if desired. For example, after installing the package (I'll cover that later in the post) I could use:

```{python}
from bitmexpy import bucket
```

To load only the `bucket` module from bitmexpy, and in order to you one of the functions I could use:

```{python}

bucket.bucket_trades().iloc[0:5, 0:6]

```


These modules can also import each other. For example, both `bucket.py` and `trades.py` have the line:

`from bitmexpy import symbols`

at the start of the file.

Here is the code for `symbols.py` as an example:

```{python}
import pandas as pd
import requests

def available_symbols():

    df = requests.get("https://www.bitmex.com/api/bitcoincharts")
    result = pd.read_json(df.content, orient="index").iloc[0]
    result = result.tolist()
    return result

AS = available_symbols()
```

`AS` is a list of all available symbols on the BitMEX exchange. This is then used in the other functions in the package by calling `symbols.AS` as a simple erro r handling technique, so if you specify the wrong symbol like:

```{python}
from bitmexpy import bucket

bucket.bucket_trades(symbol="TESLA")
```

A list of all valid symbols is given. If interested, you can check out the rest of the code in the package git repo for `bitmexpy` [here](https://github.com/hfshr/bitmexpy).

# Package metadata

For the package metadata, I found there were quite a few similarities with R packages. Below are some of the files/directories you can expect to find in a python package:

* Licence file.
* README file (Typically written in ReST or markdown).
* Changes.txt (equivalent to news.md in R).
* A `docs/` directory for package documentation.
* A `tests/` directory for any package tests (like `tests/testthat/` in R).
* Any external data that the package might rely on for examples.
* A MANIFEST.in file, describing of what non-code files to include.
* A `setup.py` file. This is very similar to the DESCRIPTION file in R and looks like this:

<aside>
By no means an exhaustive list!
</aside>

```{python eval = FALSE}
from setuptools import setup, find_packages

with open("README.rst", "r") as fh:
    long_description = fh.read()

setup(
    name="bitmexpy", 
    version="0.0.3",
    author="Harry Fisher",
    author_email="harryfisher21@gmail.com",
    description="python client for BitMEX's API",
    long_description=long_description,
    long_description_content_type="text/markdown",
    url="https://github.com/hfshr/bitmexpy",
    packages=find_packages(),
    classifiers=[
        "Programming Language :: Python :: 3",
        "License :: OSI Approved :: MIT License",
        "Operating System :: OS Independent",
    ],
    install_requires=[
        'pandas',
        'requests',
        'numpy',
        'datetime'
    ],
    python_requires=">=3.6",
)

```

As you can see from the example above, a `setup.py` file lets you:

* Outline version & package metadata
* List of packages to include
* List of other files to include
* Specify dependencies / requirements

`setup.py` is also important for building/installing your package as we'll see in the next section.

# Distribution

Aside from creating a git repository for your package, the methods of distributing packages in R and python are somewhat different. 

With a `setup.py` file, you can install the package locally by running from the package directory:

* `pip install .`

While you can now use the package locally, this isn't much help for sharing your package with others.

In order to share your package with others, the python package index (pypi) is a good choice. Conveniently, pypi has a test instance of the main index, allowing you to upload your small test packages without cluttering the main index. 

To prepare for uploading your package to pypi, you first need an account - which is easy enough to set up here: https://test.pypi.org/

Next, in the package directory run:

* `python setup.py sdist bdist_wheel`

This will create a `dist/` directory with the files like:

```{r}
fs::dir_tree("~/Downloads/Python/bitmexpy/dist/")
```

Now you can upload these files to pypi using `twine`.

* `python -m twine upload --repository testpypi dist/*`

<aside>
Will obviously need to install twine first if you haven't already!
</aside>

You will be prompted for your username and password for your testpypi account, and if all is successful you get a link to your uploaded package.

For example, you can find `bitmexpy` here: https://test.pypi.org/project/bitmexpy/0.0.5/

To install the package from pypi, you can run:

* `pip install -i https://test.pypi.org/simple/ bitmexpy==0.0.5`

Note however, not all packages on pypi are not maintained like packages on CRAN, and therefore, the quality of packages may vary considerably...! 

# Summary

So that was a quick tour of an R users attempt to understand the basics of python package building. There are certainly things I've missed here and bitmexpy needs a lot more work to get it fully functional - however python packages feel a lot less mysterious now and I have a better appreciation for how they work in general.

For further information, I found these resources useful when working on the package:

* https://python-packaging-tutorial.readthedocs.io/en/latest/setup_py.html
* https://packaging.python.org/tutorials/packaging-projects/

So definitely worth checking out if you're interested.

Thanks for reading!