---
title: "Introducing {qbr}"
description: |
  An R interface to jquery Query-Builder.
author:
  - name: Harry Fisher
    url: https://hfshr.xyz
date: 2021-06-20
preview: qbr.png
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I've been spending quite a bit of time in the world of shiny recently, and a particular problem I was facing meant I decided to dip my toe into the seemingly magical world of `htmlwidgets`. In brief, I needed a way for users to construct potential complex queries of a database. While this is certainly possible to do using native shiny inputs, I ~was being lazy~ wanted to see if I could leverage something that had already been created. After a bit of googling I came across [queryBuilder](https://querybuilder.js.org/index.html), a jquery plugin that provides UI component to create queries and filters. After playing around with the demos on the site, I was very impressed with the functionality and , and even better, someone had already made an R wrapper for the library!

However this was where I started to run into some problems. The original queryBuilder has a vast array of settings that can be configured, including additional plugins and widgets that help add additional functionality. The existing R wrappers for queryBuilder only implemented a few of these settings, and I found myself wanting to access other settings which were not yet implemented. So I thought I'd have a go at creating my own version, taking inspiration from  [harveyl888/queryBuilder](https://github.com/harveyl888/queryBuilder) with the aim of including as much of the functionality as possible of the original queryBuilder. 

## Example

Here is a quick example filtering the palmer penguins dataset with a few of the different settings configured.

<aside>
https://harryfish.shinyapps.io/qbr_demo/
</aside>

<div class="l-screen-inset">
<iframe src ="https://harryfish.shinyapps.io/qbr_demo/" allowfullscreen =""
scrolling="yes" width = "100%" height = "800px" frameboarder="0"  data-external="1"></iframe>
</div>

## Settings

The widget revolves around the list of options which you create for each filter. For example, below is the code used to specify the filters used in the demo app. 

```{r}
filters <- list(
    list(
      id = "species",
      label = "Species",
      type = "string",
      input = "select",
      description = "Shift-click to select multiple!",
      values = list("Adelie", "Gentoo", "Chinstrap"),
      multiple = TRUE,
      operators = c("equal", "not_equal", "in")
    ),
    list(
      id = "sex",
      label = "Sex",
      input = "checkbox",
      values = list(
        "male",
        "female"
      ),
      operators = c("equal", "not_equal", "in")
    ),
    list(
      id = "bill_length_mm",
      label = "Bill length",
      type = "integer",
      validation = list(
        min = 0,
        max = 100
      ),
      plugin = "slider",
      plugin_config = list(
        min = 0,
        max = 100,
        value = 0
      )
    )
  )
```

However these settings barely scratch the surface of what is possible and the original queryBuilder site (https://querybuilder.js.org/) is well worth a visit to see the full potential of the widget.

