---
title: "How has Covid affected UK athletics rankings in 2020?"
description: |
  A look into how Covid has impacted the breadth and depth of athletic performances in the UK in 2020
author:
  - name: Harry Fisher
    url: https://hfshr.xyz
date: 12-11-2020
output:
  distill::distill_article:
    self_contained: false
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
xaringanExtra::use_panelset()
```

One area that has been hugely impacted by Covid this year is sport. 


What I wanted to find out was to what extent UK athletic performance have been impacted this year, compared to other years. I have seen some incredible performances at the elite end, but I wondered if that level of athletes just below the top 2% have been affected to a greater degree.


To test my ideas I gathered some data from the UK athletics rankings website www.thepowerof10.info




```{r}
library(poweRof10)
library(tidyverse)
library(viridis)
library(furrr)
library(lubridate)
library(hms)
library(ggridges)
library(ggiraph)


### get data ###
years <- c(2016:2020)
events <- c("100", "200", "400", "800", "1500", "5000", "10000")
gender <- c("M", "W")


# grid <- expand_grid(years, events, gender) %>%
#   as.list()
# 
# plan(multisession)
# all_rankings <- future_pmap_dfr(grid, ~get_event(event = ..2, gender = ..3, year = ..1), .progress = TRUE)
# plan(sequential)

options(digits.secs=2)

load("clean_rankings.rds")

time_cleaner <- function(time) {
  date <- as.numeric(time) * 1000
  res <- as.POSIXct.numeric(date/1000, origin = '1970-01-01', format = "%OS")
  format(res, format = '%M:%OS')
}  

clean_rankings <- all_rankings %>% 
  select(rank, perf, input_year, event, gender, date, name) %>% 
  filter(event %in% c("100", "200", "400", "800", "1500", "5000")) %>% 
  drop_na(rank) %>% 
  mutate(event = factor(event, levels = c("100", "200", "400", "800", "1500", "5000")),
         rank = as.numeric(rank),
         perf = str_trim(perf),
         input_year = factor(input_year, levels = c("2016", "2017", "2018", "2019", "2020"))) %>% 
  mutate(perf = if_else(event == "400" & str_starts(perf, "6"), time_cleaner(perf), perf)) %>% 
  mutate(perf_x = case_when(event %in% c("800", "1500") ~paste0("00:0", perf),
                              event %in% c("200", "400") & nchar(perf) <= 5 ~paste0("00:00:", perf), 
                              event == "100" ~ ifelse(nchar(perf) < 5, 
                                                      paste0("00:00:0", perf), 
                                                      paste0("00:00:", perf)),
                              TRUE ~ paste0("00:", perf))) %>%  
  mutate(perf_x = parse_hms(perf_x),
         perf_x_sec = round(seconds(perf_x), digits = 2),
         perf_x_sec = round(seconds_to_period(perf_x_sec), 2)) %>% 
  mutate(test = as_datetime(perf_x_sec))



```


::::: {.panelset}

::: {.panel}
[Men]{.panel-name}


```{r fig.height=10, fig.width=10, layout = "l-page"}
library(patchwork)

df <- clean_rankings %>% 
  filter(gender == "M") %>% 
  mutate(input_year = fct_rev(input_year),
         covid_year = ifelse(input_year == "2020", "covid", "pre_covid")) %>% 
  select(input_year, perf_x, rank, event, gender, covid_year, test) %>% 
  mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20",
                                rank %in% c(21:100) ~ "20_to_100",
                                rank %in% c(101:500) ~ ">101"),
         rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))



  df %>%
  filter(rank_group == "top_20") %>% 
  ggplot(aes(x = test, y = input_year, fill = covid_year, group = input_year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_datetime(date_labels = "%M:%OS") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
  facet_wrap_custom(event~., scales = "free_x", ncol = 1, scale_overrides = list(
                      scale_override(1, scale_x_datetime(date_labels = "%OS")),
                      scale_override(2, scale_x_datetime(date_labels = "%OS")),
                      scale_override(3, scale_x_datetime(date_labels = "%OS")))) +
  labs(subtitle = "top 20") +
  theme_light() +
  theme(legend.position = "none")-> p1


df %>% 
  filter(rank_group == "20_to_100") %>% 
  ggplot(aes(x = test, y = input_year, fill = covid_year, group = input_year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_datetime(date_labels = "%M:%OS") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
    facet_wrap_custom(event~., scales = "free_x", ncol = 1, scale_overrides = list(
                      scale_override(1, scale_x_datetime(date_labels = "%OS")),
                      scale_override(2, scale_x_datetime(date_labels = "%OS")),
                      scale_override(3, scale_x_datetime(date_labels = "%OS")))) +
  labs(subtitle = "20 - 100") +
  theme_light() +
  theme(legend.position = "none",        
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) -> p2


df %>% 
  filter(rank_group == ">101") %>% 
  ggplot(aes(x = test, y = input_year, fill = covid_year, group = input_year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_datetime(date_labels = "%M:%OS") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
    facet_wrap_custom(event~., scales = "free_x", ncol = 1, scale_overrides = list(
                      scale_override(1, scale_x_datetime(date_labels = "%OS")),
                      scale_override(2, scale_x_datetime(date_labels = "%OS")),
                      scale_override(3, scale_x_datetime(date_labels = "%OS")))) +
  labs(subtitle = "Greater than 100") +
  theme_light() +
  theme(legend.position = "none",       
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) -> p3


p1 + p2 + p3 + plot_annotation(
  title = 'Distribution of performances for different rankings')
```

:::

::: {.panel}
[Women]{.panel-name}


```{r fig.height=15}
library(patchwork)

res_edit %>% 
  filter(gender == "W") %>% 
  mutate(year = fct_rev(year),
         covid_year = ifelse(year == "2020", "covid", "pre_covid")) %>% 
  select(year, perf_x, rank, event, gender, covid_year, perf_x_sec) %>% 
  mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20",
                                rank %in% c(21:100) ~ "20_to_100",
                                rank %in% c(101:max(rank)) ~ ">101"),
         rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))%>%
  filter(rank_group == "top_20") %>% 
  ggplot(aes(x = perf_x, y = year, fill = covid_year, group = year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
  facet_wrap_custom(event~, scales = "free_x", ncol = 1, 
                    scale_overrides = list(
                      scale_override(1, scale_x_continuous(breaks = c(5750, 5900))),
                      scale_override(6, scale_x_continuous(breaks = c(17800, 17900)))))
  
  
  
  facet_wrap(event ~ ., scales = "free_x", ncol = 1) +
  labs(subtitle = "top 20") +
  theme_light() +
  theme(legend.position = "none") -> p1


res_edit %>% 
  filter(gender == "W") %>% 
  mutate(year = fct_rev(year),
         covid_year = ifelse(year == "2020", "covid", "pre_covid")) %>% 
  select(year, perf_x, rank, event, gender, covid_year, perf_x_sec) %>% 
  mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20",
                                rank %in% c(21:100) ~ "20_to_100",
                                rank %in% c(101:max(rank)) ~ ">101"),
         rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))%>%
  filter(rank_group == "20_to_100") %>% 
  ggplot(aes(x = perf_x, y = year, fill = covid_year, group = year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
  facet_wrap(event ~ ., scales = "free_x", ncol = 1) +
  theme_light() +
  labs(subtitle = "20 to 100") +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) -> p2



p1 + p2
```

:::

:::::



```{r fig.height=10}
library(poweRof10)
library(tidyverse)
library(viridis)
library(furrr)
library(lubridate)
library(hms)
library(ggridges)
library(ggiraph)

#remotes::install_github("hfshr/poweRof10")

### get data ###
years <- c(2016:2020)
events <- c("100", "200", "400", "800", "1500", "5000", "10000")
gender <- c("M", "W")


grid <- expand_grid(years, events, gender) %>%
  as.list()
#
# plan(sequential)
# full_res <- future_pmap_dfr(grid, ~get_event(events = ..2, gender = ..3, year = ..1), .progress = TRUE)

years <- c(2016:2020)
events <- c("HM", "Mar")
gender <- c("M", "W")

grid <- expand_grid(years, events, gender) %>%
  as.list()

plan(multisession)
extra_res <- future_pmap_dfr(grid, ~get_event(event = ..2, gender = ..3, year = ..1, agegroup = "U17"), .progress = TRUE)


time_cleaner <- function(time) {
  date <- as.numeric(time) * 1000
  res <- as.POSIXct.numeric(date/1000, origin = '1970-01-01', format = "%OS")
  format(res, format = '%M:%OS')
}  

res_edit <- extra_res %>% 
  select(rank, perf, input_year, event, gender, date, name) %>% 
  filter(event %in% c("100", "200", "400", "800", "1500", "5000")) %>% 
  drop_na(rank) %>% 
  mutate(event = factor(event, levels = c("100", "200", "400", "800", "1500", "5000")),
         rank = as.numeric(rank),
         perf = str_trim(perf),
         input_year = factor(input_year, levels = c("2016", "2017", "2018", "2019", "2020"))) %>% 
  mutate(perf = if_else(event == "400" & str_starts(perf, "6"), time_cleaner(perf), perf)) %>% 
  mutate(perf_x = case_when(event %in% c("800", "1500") ~paste0("00:0", perf),
                              event %in% c("200", "400") & nchar(perf) <= 5 ~paste0("00:00:", perf), 
                              event == "100" ~ ifelse(nchar(perf) < 5, 
                                                      paste0("00:00:0", perf), 
                                                      paste0("00:00:", perf)),
                              TRUE ~ paste0("00:", perf))) %>%  
  mutate(perf_x = parse_hms(perf_x),
         perf_x_sec = round(seconds(perf_x), digits = 2),
         perf_x_sec = round(seconds_to_period(perf_x_sec), 2)) %>% 
  mutate(test = as_datetime(perf_x_sec))

df <- res_edit %>% 
  filter(gender == "M") %>% 
  mutate(input_year = fct_rev(input_year),
         covid_year = ifelse(input_year == "2020", "covid", "pre_covid")) %>% 
  select(input_year, perf_x, rank, event, gender, covid_year, test) %>% 
  mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20",
                                rank %in% c(21:100) ~ "20_to_100",
                                rank %in% c(101:500) ~ ">101"),
         rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))



  df %>%
  filter(rank_group == "20_to_100") %>% 
  ggplot(aes(x = test, y = input_year, fill = covid_year, group = input_year, alpha = 0.6)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_datetime(date_labels = "%M:%OS") +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
  facet_wrap_custom(event~., scales = "free_x", ncol = 1, scale_overrides = list(
                      scale_override(1, scale_x_datetime(date_labels = "%OS")),
                      scale_override(2, scale_x_datetime(date_labels = "%OS")),
                      scale_override(3, scale_x_datetime(date_labels = "%OS")))) +
  labs(subtitle = "top 20") +
  theme_light() +
  theme(legend.position = "none")









full_res <- read.csv("raw_data.csv")


  mutate(year = as.character(year),
         event = as.character(event)) %>%
  bind_rows(., extra_res %>%
              select(rank, perf, year = year_x, event, gender, date, name) %>%
              mutate(rank = as.numeric(rank))) %>%
  drop_na(rank)

test_hm <- full_res %>%
  filter(event == "HM") %>%
  mutate()

```




<!-- res_edit <- full_res %>%  -->
<!--   filter(rank != "") %>%  -->
<!--   mutate(event = factor(event, levels = c("100", "200", "400", "800", "1500", "5000", "10000", "HM", "Mar"))) %>%  -->
<!--   mutate(rank = as.numeric(rank), -->
<!--          perf = str_trim(perf)) %>%  -->
<!--     mutate(perf_x = case_when(event %in% c("800", "1500") ~paste0("00:0", perf), -->
<!--                               event %in% c("200", "400") ~ ifelse(nchar(perf) <= 5, paste0("00:00:", perf), paste0("00:0", perf)), -->
<!--                               event %in% c("HM", "Mar") ~ ifelse(nchar(perf) <= 5, paste0("00:", perf), paste0("0", perf)), -->
<!--                               event %in% c("5000", "10000") ~ ifelse(nchar(perf) <= 8, paste0("00:", perf), paste0("0", perf)), -->
<!--                               event == "100" ~ ifelse(nchar(perf) < 5, paste0("00:00:0", perf), paste0("00:00:", perf)), -->
<!--                               TRUE ~ perf)) %>%      -->
<!--   mutate(perf_x = parse_hms(perf_x), -->
<!--          perf_x_sec = round(seconds(perf_x), digits = 2), -->
<!--          year = factor(year, levels = c("2016", "2017", "2018", "2019", "2020")))  -->


<!-- res_edit <- full_res %>%  -->
<!--   select(rank, perf, year, event, gender, date, name) %>%  -->
<!--   filter(event %in% c("100", "200", "400", "800", "1500", "5000")) %>%  -->
<!--   filter(rank != "") %>%  -->
<!--   mutate(event = factor(event, levels = c("100", "200", "400", "800", "1500", "5000"))) %>%  -->
<!--   mutate(rank = as.numeric(rank), -->
<!--          perf = str_trim(perf)) %>%  -->
<!--     mutate(perf_x = case_when(event %in% c("800", "1500") ~paste0("00:0", perf), -->
<!--                               event %in% c("200", "400") ~ ifelse(nchar(perf) <= 5, paste0("00:00:", perf), paste0("00:0", perf)), -->
<!--                               event == "100" ~ ifelse(nchar(perf) < 5, paste0("00:00:0", perf), paste0("00:00:", perf)), -->
<!--                               TRUE ~ paste("00:", perf)))%>%      -->
<!--   mutate(perf_x = parse_hms(perf_x), -->
<!--          perf_x_sec = round(seconds(perf_x), digits = 2), -->
<!--          year = factor(year, levels = c("2016", "2017", "2018", "2019", "2020"))) %>%  -->
<!--   mutate(perf_x_sec = seconds_to_period(perf_x_sec)) -->


<!-- library(ggrepel) -->

<!-- df <- res_edit %>%  -->
<!--   filter(year %in% c("2020","2019")) %>%  -->
<!--   mutate(year = factor(year, levels = c("2019", "2020"))) %>%  -->
<!--   group_by(year, event, gender) %>%  -->
<!--   summarise(n = n())  -->



<!-- df %>%  -->
<!--   ggplot(aes(x = year, y = n, label = event, colour = event)) + -->
<!--   geom_point(size = 3) + -->
<!--   geom_line(aes(group = event), size =2) + -->
<!--   scale_color_brewer(palette="Set2") + -->
<!--   geom_label_repel( -->
<!--     data = subset(df, year == "2019"), -->
<!--     nudge_x      = -0.20, -->
<!--     direction    = "y", -->
<!--     hjust        = 1, -->
<!--     segment.size = 0.2) + -->
<!--   geom_point_interactive(aes(tooltip = paste0("n = ", n)), size = 2) + -->
<!--   facet_wrap(~gender) + -->
<!--   theme_minimal() + -->
<!--   theme(legend.position = "none") -> my_gg -->


<!--   girafe(code = print(my_gg) ,height_svg = 4) -->

<!-- ``` -->


<!-- # Breakdown by rankings -->

<!-- ## top 10 -->


<!-- ```{r fig.height=15} -->
<!-- library(patchwork) -->

<!-- res_edit %>%  -->
<!--   filter(gender == "M") %>%  -->
<!--   mutate(year = fct_rev(year), -->
<!--          covid_year = ifelse(year == "2020", "covid", "pre_covid")) %>%  -->
<!--   select(year, perf_x, rank, event, gender, covid_year, perf_x_sec) %>%  -->
<!--   mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20", -->
<!--                                 rank %in% c(21:100) ~ "20_to_100", -->
<!--                                 rank %in% c(101:max(rank)) ~ ">101"), -->
<!--          rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))%>% -->
<!--   filter(rank_group == "top_20") %>%  -->
<!--   ggplot(aes(x = perf_x, y = year, fill = covid_year, group = year, alpha = 0.6)) + -->
<!--   scale_fill_brewer(palette = "Set1") + -->
<!--   stat_density_ridges(quantile_lines = TRUE, quantiles = 2) + -->
<!--   facet_wrap(event ~ ., scales = "free_x", ncol = 1) + -->
<!--   labs(subtitle = "top 20") + -->
<!--   theme_light() + -->
<!--   theme(legend.position = "none") -> p1 -->


<!-- res_edit %>%  -->
<!--   filter(gender == "M") %>%  -->
<!--   mutate(year = fct_rev(year), -->
<!--          covid_year = ifelse(year == "2020", "covid", "pre_covid")) %>%  -->
<!--   select(year, perf_x, rank, event, gender, covid_year, perf_x_sec) %>%  -->
<!--   mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20", -->
<!--                                 rank %in% c(21:100) ~ "20_to_100", -->
<!--                                 rank %in% c(101:max(rank)) ~ ">101"), -->
<!--          rank_group = factor(rank_group, levels = c("top_20", "20_to_100", ">101")))%>% -->
<!--   filter(rank_group == "20_to_100") %>%  -->
<!--   ggplot(aes(x = perf_x, y = year, fill = covid_year, group = year, alpha = 0.6)) + -->
<!--   scale_fill_brewer(palette = "Set1") + -->
<!--   stat_density_ridges(quantile_lines = TRUE, quantiles = 2) + -->
<!--   facet_wrap(event ~ ., scales = "free_x", ncol = 1) + -->
<!--   theme_light() + -->
<!--   labs(subtitle = "20 to 100") + -->
<!--   theme(legend.position = "none", -->
<!--         axis.text.y = element_blank(), -->
<!--         axis.title.y = element_blank()) -> p2 -->



<!-- p1 + p2 -->
<!-- ``` -->

## 21 to 100

```{r}
res_edit %>% 
  filter(gender == "M") %>% 
  mutate(year = fct_rev(year),
         covid_year = ifelse(year == "2020", "covid", "pre_covid")) %>% 
  select(year, perf_x, rank, event, gender, covid_year) %>% 
  mutate(rank_group = case_when(rank %in% c(1:20) ~ "top_20",
                                rank %in% c(21:100) ~ "20_to_100",
                                rank %in% c(101:250) ~ "101_to_250",
                                TRUE ~ "NA"),
         rank_group = factor(rank_group, levels = c("top_20", "20_to_100", "101_to_250"))) %>%
  filter(rank_group != "NA") %>% 
  filter(rank_group == "top_20") %>% 
  ggplot(aes(x = perf_x, y = year, fill = covid_year, group = year)) +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2) +
  facet_wrap(event ~., dir = "v", scales = "free_x")

```

# test

<!-- ```{r} -->
<!-- library(ggalt) -->

<!-- df <- res_edit %>%  -->
<!--   filter(year %in% c("2020", "2019")) %>%  -->
<!--   mutate(covid_year = ifelse(year == "2020", "covid", "non_covid"), -->
<!--          covid_year = factor(covid_year, levels = c("non_covid", "covid"))) %>% -->
<!--   group_by(covid_year, event, gender) %>%  -->
<!--   summarise(n = n())%>%  -->
<!--   mutate(across(.cols = n, round, 0)) %>%  -->
<!--   pivot_wider(values_from = n, names_from = covid_year) %>%  -->
<!--   mutate(perc_dec = round((non_covid - covid)/non_covid * 100, digits = 0)) %>%  -->
<!--   mutate(perc_dec = paste0("-", perc_dec, "%")) %>%  -->
<!--   arrange(-non_covid) %>%  -->
<!--   filter(gender == "M") -->

<!-- library(ggtext) -->
<!-- library(ggalt) -->

<!-- ggplot() + -->
<!--   geom_segment(data=df, aes(y=event, yend=event, x=0, xend=1)) + -->
<!--   geom_dumbbell(data=df, aes(y=event, x=covid, xend=non_covid), -->
<!--                 color="#b2b2b3",size = 4,colour_x="#9fb059", colour_xend="#edae52")+ -->
<!--   geom_text(data=filter(df, event == "100"), -->
<!--                      aes(x=covid, y=event, label="2020 ðŸ¦ "), -->
<!--                      color="#9fb059", size=4, vjust=3.2, fontface="bold", family="Calibri")+  -->
<!--   geom_text(data=filter(df, event == "100"), -->
<!--                      aes(x=non_covid, y=event, label="2019 ðŸƒ"), -->
<!--                      color="#edae52", size=4, vjust=-3, fontface="bold", family="Calibri")   + -->
<!--   # geom_text(data=filter(df, event == "100"), -->
<!--   #                    aes(x=(non_covid + covid) / 2, y=event, label="ðŸ”»"), -->
<!--   #                    color="red", size=4, hjust=1.1, vjust = -2, fontface="bold", family="Calibri")   + -->
<!--   geom_text(data=df, aes(x=covid, y=event, label=covid), -->
<!--                      color="#9fb059", size=3.5, vjust=2, family="Calibri") +  -->
<!--   geom_text(data=df, color="#edae52", size=3.5, vjust=-1.5, family="Calibri", -->
<!--                      aes(x=non_covid, y=event, label=non_covid)) + -->
<!--   geom_text(data = df, color="red", size=3.5, hjust=1.2, family="Calibri", -->
<!--             aes(x = (non_covid + covid) / 2 , y = event, label = perc_dec)) + -->
<!--   scale_x_continuous(limits = c(-10, 1600)) + -->
<!--   coord_flip() + -->
<!--   scale_y_discrete(expand=c(0.075,0.6)) + -->
<!--   labs(y = "Event", x = "Number of athletes on ranking list", -->
<!--        title = "<span style = 'color:red;'>Percentage decrease</span>  -->
<!--        in the number of athletes on the ranking \nlist for  -->
<!--        <span style = 'color:#9fb059;'>2020 (covid)</span>  -->
<!--        vs   <span style = 'color:#edae52;'>2019 (pre-covid)</span>", -->
<!--        subtitle = "") + -->
<!--   theme_minimal() + -->
<!--   theme(plot.title = element_textbox_simple(size = 16), -->
<!--         axis.text.y = element_text(size = 14, face = "bold", family = "Calibri"), -->
<!--         axis.text.x = element_text(size = 14, face = "bold", family = "Calibri"), -->
<!--         axis.title = element_blank())  -->



<!-- library(poweRof10) -->
<!-- library(tidyverse) -->


<!-- ``` -->


<!-- https://fishandwhistle.net/post/2018/modifying-facet-scales-in-ggplot2/ -->

```{r }
scale_override <- function(which, scale) {
  if(!is.numeric(which) || (length(which) != 1) || (which %% 1 != 0)) {
    stop("which must be an integer of length 1")
  }
  
  if(is.null(scale$aesthetics) || !any(c("x", "y") %in% scale$aesthetics)) {
    stop("scale must be an x or y position scale")
  }
  
  structure(list(which = which, scale = scale), class = "scale_override")
}

CustomFacetWrap <- ggproto(
  "CustomFacetWrap", FacetWrap,
  init_scales = function(self, layout, x_scale = NULL, y_scale = NULL, params) {
    # make the initial x, y scales list
    scales <- ggproto_parent(FacetWrap, self)$init_scales(layout, x_scale, y_scale, params)
    
    if(is.null(params$scale_overrides)) return(scales)
    
    max_scale_x <- length(scales$x)
    max_scale_y <- length(scales$y)
    
    # ... do some modification of the scales$x and scales$y here based on params$scale_overrides
    for(scale_override in params$scale_overrides) {
      which <- scale_override$which
      scale <- scale_override$scale
      
      if("x" %in% scale$aesthetics) {
        if(!is.null(scales$x)) {
          if(which < 0 || which > max_scale_x) stop("Invalid index of x scale: ", which)
          scales$x[[which]] <- scale$clone()
        }
      } else if("y" %in% scale$aesthetics) {
        if(!is.null(scales$y)) {
          if(which < 0 || which > max_scale_y) stop("Invalid index of y scale: ", which)
          scales$y[[which]] <- scale$clone()
        }
      } else {
        stop("Invalid scale")
      }
    }
    
    # return scales
    scales
  }
)

facet_wrap_custom <- function(..., scale_overrides = NULL) {
  # take advantage of the sanitizing that happens in facet_wrap
  facet_super <- facet_wrap(...)
  
  # sanitize scale overrides
  if(inherits(scale_overrides, "scale_override")) {
    scale_overrides <- list(scale_overrides)
  } else if(!is.list(scale_overrides) || 
            !all(vapply(scale_overrides, inherits, "scale_override", FUN.VALUE = logical(1)))) {
    stop("scale_overrides must be a scale_override object or a list of scale_override objects")
  }
  
  facet_super$params$scale_overrides <- scale_overrides
  
  ggproto(NULL, CustomFacetWrap,
    shrink = facet_super$shrink,
    params = facet_super$params
  )
}

```


