---
title: "Pretty tables with gt"
description: |
  Traditionally I have been an ardent user of kable + kableExtra when it comes to creating tables. These packages have served me well, however the CRAN release of a new player in the table package space - gt - promted me to try it out and explore some of the features it had to offer.
author:
  - name: Harry Fisher
    url: https://hfshr.netlify.com/
date: 05-02-2020
output:
  distill::distill_article:
    self_contained: false
draft: false
preview: nice_table.png
categories:
  - Bitcoin
  - gt
  - ggplot
  - bitmexr
  - R
bibliography: r-references.bib  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(bitmexr)
library(gt)
library(tidyverse)
library(tidyquant)
papaja::r_refs(file = "r-references.bib")
```

`gt` [@R-gt] has been under development for a while, but a stable version has recently been released to CRAN. In order to try the package out we first need some data for creating tables.
I'll be using some Bitcoin price data (shameless self plug for my `bitmexr` [@R-bitmexr] package that has just landed on CRAN - check it out [here](https://cran.r-project.org/web/packages/bitmexr/index.html)).

For demonstration purposes, I pulled the OHLC prices for the last 10 days.

```{r}
bitcoin <- bucket_trades(
  binSize = "1d",
  startTime = "2020-04-23",
  endTime = "2020-05-02"
) %>%
  select(1:8) %>%
  select(-symbol)


head(bitcoin)
```

The most basic table without any formatting applied looks like this:

```{r}
bitcoin %>%
  gt()
```


To get started, I borrowed some ideas straight from an example on https://gt.rstudio.com/ and applied some simple formatting to the raw data in the table. 

- Added a header containing the title and subtitle of the table
- Format the price columns with the appropriate currency symbol
- Format the date column
- Format the trades and volume column with appropriate suffixes

```{r}
start_date <- format(min(bitcoin$timestamp), "%d/%m/%Y")
end_date <- format(max(bitcoin$timestamp), "%d/%m/%Y")

tab <- bitcoin %>%
  mutate(timestamp = as.character(timestamp)) %>%
  gt() %>%
  tab_header(
    title = md("**Bitcoin price**"),
    subtitle = glue::glue("{start_date} to {end_date}")
  ) %>%
  fmt_currency(
    columns = vars(open, high, low, close),
    currency = "USD"
  ) %>%
  fmt_number(
    columns = vars(trades, volume),
    suffixing = TRUE
  ) %>%
  fmt_date(
    columns = vars(timestamp),
    date_style = 9
  )

tab
```


Starting to look a little more presentable. Next, I

- Added some colour to highlight changes in the trades and volume columns
- Applied some formatting to the column names.
- I also changed the number of decimal places to 0 (2 decimal places seemd overkill for this example).

A cool feature of `gt` is that you can rename the column labels to something different, but refer to the original labels for any further manipulation. 


```{r}
tab <- tab %>%
  tab_spanner(
    label = "Price",
    columns = vars(open, high, low, close)
  ) %>%
  data_color(
    columns = vars(trades, volume),
    colors = scales::col_numeric("Greens", domain = NULL)
  ) %>%
  cols_align(
    align = "center",
    columns = vars(trades, volume)
  ) %>%
  cols_label(
    timestamp = "Date",
    open = md("**Open**"),
    high = "High",
    low = "Low",
    close = md("**Close**"),
    trades = "Trades",
    volume = "Volume"
  ) %>%
  fmt_currency(
    columns = vars(open, high, low, close),
    currency = "USD",
    decimals = 0
  )


tab
```




Finally, it is also very easy to add custom html to the tables using the `text_transformations()` function (again inspired by https://gt.rstudio.com/). 
The rocket and sad face indicate whether the closing price was greater or less than the opening price for each dat

```{r}
rocket <- "<span style=\"color:green\">&#128640;</span>"
sadface <- "<span style=\"color:red\">&#128546;</span>"

tab <- tab %>%
  text_transform(
    locations = cells_body(
      columns = "close",
      rows = close > open
    ),
    fn = function(x) paste(x, rocket)
  ) %>%
  text_transform(
    locations = cells_body(
      columns = "close",
      rows = close < open
    ),
    fn = function(x) paste(x, sadface)
  )

tab
```

Looks good! I've barely scratched the surface of what `gt` can do here and the [package website](https://gt.rstudio.com/) is well worth a look if you are interested.  

## Bonus table

One experimental feature in `gt` is the ability to add a ggplot (or any image for that matter) to a table. Yes thats right, a ggplot in a table. I did have to do a bit of rummaging in the package repo to get this to work, but the idea is really cool.

To demonstrate this functionality I pulled some new Bitcoin price data - the hourly OHLC for the whole on April 2020. The goal is to summarise this information for weekly intervals, with the total volume and trades and a plot of the price action for that week.

First I define a few functions that will help in creating the table/plot combo.

```{r}
# helper for summary data
oc <- function(name, df) {
  df %>%
    summarise(
      open = dplyr::first(open),
      close = dplyr::last(close),
      volume = sum(volume),
      trades = sum(trades),
      date = dplyr::first(timestamp)
    )
}

# ggplot for each week
plot_group <- function(name, df) {
  plot_object <-
    ggplot(
      data = df,
      aes(x = timestamp, y = close)
    ) +
    geom_barchart(
      aes(open = open, high = high, low = low, close = close),
      fill_up = "green",
      fill_down = "red",
      colour_up = "green",
      colour_down = "red"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.title = element_blank(),
      axis.text = element_blank()
    )
  return(plot_object)
}

# combine ggplot within table
fmt_ggplot <- fmt_gg <- function(
  data,
  columns,
  rows = NULL,
  height = 100,
  aspect_ratio = 1.0
) {
  rows <- rlang::enquo(rows)

  fmt(
    data = data,
    columns = columns,
    rows = !!rows,
    fns = list(
      html = function(x) {
        map(
          x,
          ggplot_image,
          height = height,
          aspect_ratio = aspect_ratio
        )
      }
    )
  )
}
```

1. The first is just a simple helper function which generates some summary information for the final table

2. The second is the ggplot object that will be used to visualise each weeks data.

3. The third was borrowed straight from gt's repo [here](https://github.com/rstudio/gt/pull/155/commits/fb862bdd16a8c7236727e51ddeff700cd2540b75). I think some of this functionality is still under development, so is not yet available in the CRAN release.

We're now ready to prepare the data for the table.

```{r}
newdata <- map_bucket_trades(
  start_date = "2020-04-01",
  end_date = "2020-04-28",
  binSize = "1h"
) %>%
  mutate(week = lubridate::week(timestamp)) %>%
  select(week, timestamp, open, high, low, close, volume, trades) %>%
  group_by(week) %>%
  nest() %>%
  mutate(
    plot = map2(week, data, plot_group),
    info = map2(week, data, oc)
  ) %>%
  unnest(info)

head(newdata)
```


Now the data is ready, we can make the table. 

```{r}

newdata %>%
  ungroup() %>%
  select(date, volume, trades, open, close, plot) %>%
  mutate(date = as.character(date)) %>%
  gt() %>%
  fmt_ggplot(
    columns = vars(plot),
    height = 170,
    aspect_ratio = 1
  ) %>%
  fmt_number(
    columns = vars(trades, volume),
    suffixing = TRUE
  ) %>%
  fmt_date(
    columns = vars(date),
    date_style = 8
  ) %>%
  cols_align(
    align = "center"
  ) %>%
  fmt_currency(
    columns = vars(open, close)
  ) %>%
  tab_header(
    title = md("**Price action summary for April 2020**")
  ) %>%
  cols_label(
    date = "Week beginning",
    open = "Open",
    close = "Close",
    trades = "Trades",
    volume = "Volume",
    plot = "Price action"
  ) %>%
  tab_spanner(
    label = "Summary",
    columns = vars(volume, trades, open, close)
  )
```


Pretty cool! The possibilities are endless with this and it was good fun trying to get it to work. 
