---
name: Harry Fisher
position: "Academic Associate"
address: "11 Thorley Cardiff"
phone: "07716891305"
email: "harryfisher21@gmail.com"
github: hfshr
date: "`r format(Sys.time(), '%B %Y')`"
headcolor: "000088"
output: 
  vitae::hyndman:
    keep_tex: TRUE
linestretch: 1.25
header-includes:
  - \usepackage[none]{hyphenat}
  - \usepackage{ragged2e}
bibliography: refs.bib
nocite: |
  @Fisher2016
citation_package: natbib
urlcolor: blue
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(vitae)
library(rorcid)
library(scholar)
library(tidyverse)
```

# Profile

\justify

About me - Currently a statistical modeller

# Key skills and Attributes

__Data science | Predictive modelling | R | Python | Spark | Interactive dashboards | Adaptable | Driven | Problem solving__

# Education

```{r education}

tribble(
  ~year, ~degree, ~institution,
  "2016 - 2019", "Ph.D.", "Cardiff Metropolitan University",
  "2013 - 2015", "M.Sc", "University of Wyoming",
  "2011 - 2013", "B.Sc.(Hons)", "Cardiff Metropolitan University"
  ) %>%
  detailed_entries(degree, year, where = institution)
```


# Skills developed from Ph.D.

- Substantial experience using R and R Markdown (self taught) to run complex statistical models and produce high quality technical documents throughout my Ph.D.
- Proficient in a number of data cleaning, wrangling and visualisation techniques using a wide variety of methods in R.
- Strong statistical knowledge using several different techniques including Bayesian multilevel modelling, generalised linear models and probabilistic network analysis.
- Experience using the CARET (Kuhn et al., 2018) package in R for fitting random forests and naive Bayes classifiers as part of exploratory data analysis for my Ph.D. results.
- Basic understanding of key libraries in python including NumPy, Pandas and BeautifulSoup as well as using R package sparklyr for accessing Apache Spark to work with large data sets.
- Experience using GIT and Github for my own version control during my Ph.D. write-up and for sharing reports with my supervisory team.
- Collaboration with other Ph.D. students to work on research projects involving large datasets of rugby injuries, requiring a high level of communication and effective team work to ensure timely data analysis and dissemination of results.
- Strong ability to plan and work independently, while remaining flexible to honour other commitments such as coaching and voluntary work.

# Employment

```{r employment}

tribble(
  ~year, ~role, ~loc,
  "2016 - Current", "Academic Associate", "Cardiff Metropolitan University",
  "2017-2018", "Sports Science Support", "Sport Wales (Voluntary role)",
  "2015 - 2016", "Athletics Coach", "Cardiff Metropolitan University"
) %>% 
  detailed_entries(year, what = role, where = loc)

```


# Skills developed from employment

- My role as an academic associate included:

    - Planning and delivery of lead lectures and seminars for first and second year students.
    - Administrative roles including marking coursework and examinations.
    - Collaboration with the module leader to ensure high standards of teaching were met, maintaining high levels of student satisfaction on the modules which I was responsible for.
    
- My voluntary role with Sport Wales included: 

     - Collaboration with several members of the sport science support team to record, analyse and disseminate data to relevant parties on athletes' performance and injury status in a timely manner. 
     - Critically review the information that was available and use it to inform future training or competition. 
     - Tailor the delivery of results to suit the audience (e.g., some coaches preferred a simple report with only the key findings, whereas others preferred a detailed analysis of all relevant data). 

# Achievements

- Secured funding for my Ph.D. through the highly competitive Research Innovation Award at Cardiff Metropolitan University (2016).

# Projects on Github

- Interactive dashboard for visualising athletics results: \textcolor{blue}{https://bit.ly/2KbpTdu}
- Summary of Ph.D. study using Bayesian Networks:\textcolor{blue}{https://bit.ly/31aqkM1}
- Discussion of analysis techniques considered in Ph.D.:\textcolor{blue}{https://bit.ly/316nDuy}

# Publications 
```{r}
scholar::get_publications("4_dgVaIAAAAJ&hl")%>% 
  head(1) %>% 
  detailed_entries(
    what = title,
    when = year,
    with = author,
    where = glue::glue("{`journal`}, {`number`}")
  )
```

# References

_References are available upon request_

